Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-19 docs/manuscript.Rmd
updated

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$). We then went on to predict the causal directions of the associations for varying levels of possible measurement error. `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test. We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b). There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`). Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.

@@

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation. Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest. But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation in have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages. First, it uses a formal statistical framework to test for the robustness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

@@

In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available. However, MR strategies exist for other scenarios, for example if there are multiple instruments for the hypothesised exposure then the problems described here may be circumvented [@DaveySmithHemani2014]. Likewise, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014]. In this work we assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored. An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis. However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood. We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies. We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated, however our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.

@@

The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is highly sensitive to measurement error. With the presence of measurement error near ubiquitous in most observational data, and our ability to measure it limited, we argue that it needs to be central to any consideration of approaches which are used in attempt to strengthen causal inference, and any putative results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.

>>>>

We used the Steiger test to infer the direction of causality between `r nrow(qtl_orig)` putative associations between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error, DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$). The median reliability (R) of the `r nrow(qtl_orig)` tests was `r round(shakh_median_reliability, 2)` (95% CI `r round(shakh_r_loci, 2)` - `r round(shakh_r_upci, 2)`). We then went on to predict the causal directions of the associations for varying levels of systematic measurement error for the different platforms. `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

We performed two sample MR [@Pierce2013] for each association in the direction of causality inferred by the Stieger test. We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b). There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`). Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.

@@

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation. Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest. But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages. First, it uses a formal statistical framework to test for the robustness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

@@

In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available. However, MR strategies exist for other scenarios, for example if there are multiple instruments for the hypothesised exposure then the problems described here may be circumvented [@DaveySmithHemani2014]. Likewise, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014]. We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].

In this work we assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce when multiple instruments are available, but how they perform in the presence of measurement error remains to be explored. An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis. However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood. We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies. We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated, however our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.

@@

The overarching result from our simulations is that, regardless of the method used, inferring the causal direction using an instrument of unknown biology is highly sensitive to measurement error. With the presence of measurement error near ubiquitous in most observational data, and our ability to measure it limited, we argue that it needs to be central to any consideration of approaches which are used in attempt to strengthen causal inference, and any putative results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.

### 2017-02-19 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{18 February 2017}

@@

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels and found that the causal
direction commonly goes in both directions (Figure 6a), but assuming no
or equal measurement error, DNA methylation levels were the predominant
causal factor (\(p = 1.3\times 10^{-5}\)). We then went on to predict
the causal directions of the associations for varying levels of possible
measurement error. Figure 6a shows that the qualitative result of which
of DNA methylation or gene expression is more commonly the causal factor
is very strongly determined by measurement error.

We performed two sample MR\textsuperscript{28} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
generally in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{40} (Figure 6b). There was
a moderate correlation between the absolute magnitudes of the causal
correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even for relatively low level
associations it is important to use causal inference methods over
observational associations to infer causal effect sizes.

@@

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of methods such as the CIT or similar
that are based on mediation. Inferring the existence of causality using
Mendelian randomisation is robust in the face of measurement error and,
if the researcher has knowledge about the biology of the instrument
being used in the analysis, can offer a direct solution to the issues
that the CIT faces. This assumption is often reasonable, for example
SNPs are commonly used as instruments when they are found in genes with
known biological relevance for the trait of interest. But on many
occasions, especially in the realm of 'omic data, this is not the case,
and methods based on mediation in have been valuable in order to be able
to both ascertain if there is a causal association and to infer the
direction of causality. Here we have described a simple extension to MR
which can be used as an alternative to or in conjunction with mediation
based methods. We show that this method is still liable to measurement
error, but because it has different properties to the CIT it offers
several main advantages. First, it uses a formal statistical framework
to test for the robustness of the assumed direction of causality.
Second, after testing in a comprehensive range of scenarios the MR based
approach is less likely to infer the wrong direction of causality
compared to CIT, while substantially improving power over CIT in the
cases where \(d > 0\).

@@

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
MR strategies exist for other scenarios, for example if there are
multiple instruments for the hypothesised exposure then the problems
described here may be circumvented\textsuperscript{16}. Likewise, if
there is at least one instrument for each trait then bi-directional MR
can offer solutions to inferring the causal
direction\textsuperscript{46}. In this work we assumed that pleiotropy
(the influence of the instrument on the outcome through a mechanism
other than the exposure) was not present. Recent method developments in
MR\textsuperscript{24,25} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established\textsuperscript{34} and have a
number of extensions that make them valuable tools, including for
example network construction. But because they are predicated on the
basic underlying principles of mediation they are liable to suffer from
the same issues of measurement error. Recent advances in MR methodology,
for example applying MR to genetical genomics\textsuperscript{47},
multivariate MR\textsuperscript{48} and mediation through
MR\textsuperscript{49--51} may offer more robust alternatives for these
more complicated problems.

The overarching result from our simulations is that, regardless of the
method used, inferring causal direction using an instrument of unknown
biology is highly sensitive to measurement error. With the presence of
measurement error near ubiquitous in most observational data, and our
ability to measure it limited, we argue that it needs to be central to
any consideration of approaches which are used in attempt to strengthen
causal inference, and any putative results should be accompanied with
appropriate sensitivity analysis that assesses their robustness under
varying levels of measurement error.

@@

\hypertarget{ref-Relton2012}{}
47. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
randomization: a strategy for establishing the causal role of epigenetic
processes in pathways to disease. \emph{International journal of
epidemiology} \textbf{41,} 161--76 (2012).

\hypertarget{ref-Burgess2014a}{}
48. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
G. Using multivariable Mendelian randomization to disentangle the causal
effects of lipid fractions. \emph{PloS one} \textbf{9,} e108891 (2014).

\hypertarget{ref-Varbo2015}{}
49. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).

\hypertarget{ref-Burgess2015}{}
50. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
Network Mendelian randomization: using genetic variants as instrumental
variables to investigate mediation in causal pathways.
\emph{International journal of epidemiology} \textbf{44,} 484--95
(2015).

\hypertarget{ref-Richmond2016}{}
51. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
C. L. Challenges and novel approaches for investigating molecular
mediation. \emph{Human molecular genetics} \textbf{25,} R149--R156
(2016).

>>>>

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{19 February 2017}

@@

We used the Steiger test to infer the direction of causality between 458
putative associations between DNA methylation levels and gene expression
levels and found that the causal direction commonly goes in both
directions (Figure 6a), but assuming no or equal measurement error, DNA
methylation levels were the predominant causal factor
(\(p = 1.3\times 10^{-5}\)). The median reliability (R) of the 458 tests
was 3.92 (95\% CI 1.08 - 37.11). We then went on to predict the causal
directions of the associations for varying levels of systematic
measurement error for the different platforms. Figure 6a shows that the
qualitative result of which of DNA methylation or gene expression is
more commonly the causal factor is very strongly determined by
measurement error.

We performed two sample MR\textsuperscript{28} for each association in
the direction of causality inferred by the Stieger test. We observed
that the sign of the MR estimate was generally in the same direction as
the Pearson correlation coefficient reported by Shakhbazov et
al\textsuperscript{40} (Figure 6b). There was a moderate correlation
between the absolute magnitudes of the causal correlation and the
observational Pearson correlation (r = 0.45). Together these inferences
suggest that even for relatively low level associations it is important
to use causal inference methods over observational associations to infer
causal effect sizes.

@@

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of methods such as the CIT or similar
that are based on mediation. Inferring the existence of causality using
Mendelian randomisation is robust in the face of measurement error and,
if the researcher has knowledge about the biology of the instrument
being used in the analysis, can offer a direct solution to the issues
that the CIT faces. This assumption is often reasonable, for example
SNPs are commonly used as instruments when they are found in genes with
known biological relevance for the trait of interest. But on many
occasions, especially in the realm of 'omic data, this is not the case,
and methods based on mediation have been valuable in order to be able to
both ascertain if there is a causal association and to infer the
direction of causality. Here we have described a simple extension to MR
which can be used as an alternative to or in conjunction with mediation
based methods. We show that this method is still liable to measurement
error, but because it has different properties to the CIT it offers
several main advantages. First, it uses a formal statistical framework
to test for the robustness of the assumed direction of causality.
Second, after testing in a comprehensive range of scenarios the MR based
approach is less likely to infer the wrong direction of causality
compared to CIT, while substantially improving power over CIT in the
cases where \(d > 0\).

@@

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
MR strategies exist for other scenarios, for example if there are
multiple instruments for the hypothesised exposure then the problems
described here may be circumvented\textsuperscript{16}. Likewise, if
there is at least one instrument for each trait then bi-directional MR
can offer solutions to inferring the causal
direction\textsuperscript{46}. We restricted the simulations to
evaluating the causal inference between quantitative traits, but it is
possible that the analysis could be extended to binary traits by using
the genetic variance explained on the liability scale, taking into
account the population prevalence\textsuperscript{47}.

In this work we assumed that pleiotropy (the influence of the instrument
on the outcome through a mechanism other than the exposure) was not
present. Recent method developments in MR\textsuperscript{24,25} have
focused on accounting for the issues that horizontal pleiotropy can
introduce when multiple instruments are available, but how they perform
in the presence of measurement error remains to be explored. An
important advantage that MR confers over most mediation based analysis
is that it can be performed in two samples, which can considerably
improve power and expand the scope of analysis. However, whether there
is a substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established\textsuperscript{34} and have a
number of extensions that make them valuable tools, including for
example network construction. But because they are predicated on the
basic underlying principles of mediation they are liable to suffer from
the same issues of measurement error. Recent advances in MR methodology,
for example applying MR to genetical genomics\textsuperscript{48},
multivariate MR\textsuperscript{49} and mediation through
MR\textsuperscript{50--52} may offer more robust alternatives for these
more complicated problems.

The overarching result from our simulations is that, regardless of the
method used, inferring the causal direction using an instrument of
unknown biology is highly sensitive to measurement error. With the
presence of measurement error near ubiquitous in most observational
data, and our ability to measure it limited, we argue that it needs to
be central to any consideration of approaches which are used in attempt
to strengthen causal inference, and any putative results should be
accompanied with appropriate sensitivity analysis that assesses their
robustness under varying levels of measurement error.

@@

\hypertarget{ref-Lee2013c}{}
47. Lee, S. H. \& Wray, N. R. Novel genetic analysis for case-control
genome-wide association studies: quantification of power and genomic
prediction accuracy. \emph{PLoS One} \textbf{8,} e71494 (2013).

\hypertarget{ref-Relton2012}{}
48. Relton, C. L. \& Davey Smith, G. Two-step epigenetic Mendelian
randomization: a strategy for establishing the causal role of epigenetic
processes in pathways to disease. \emph{International journal of
epidemiology} \textbf{41,} 161--76 (2012).

\hypertarget{ref-Burgess2014a}{}
49. Burgess, S., Freitag, D. F., Khan, H., Gorman, D. N. \& Thompson, S.
G. Using multivariable Mendelian randomization to disentangle the causal
effects of lipid fractions. \emph{PloS one} \textbf{9,} e108891 (2014).

\hypertarget{ref-Varbo2015}{}
50. Varbo, A. \emph{et al.} Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. \emph{Circulation research} \textbf{116,} 665--73 (2015).

\hypertarget{ref-Burgess2015}{}
51. Burgess, S., Daniel, R. M., Butterworth, A. S. \& Thompson, S. G.
Network Mendelian randomization: using genetic variants as instrumental
variables to investigate mediation in causal pathways.
\emph{International journal of epidemiology} \textbf{44,} 484--95
(2015).

\hypertarget{ref-Richmond2016}{}
52. Richmond, R. C., Hemani, G., Tilling, K., Davey Smith, G. \& Relton,
C. L. Challenges and novel approaches for investigating molecular
mediation. \emph{Human molecular genetics} \textbf{25,} R149--R156
(2016).
